```{r}

```

---
title: "Final Project"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

# Final Project: JFK Passengers

**Source.**
<https://github.com/alan-turing-institute/TCPD/tree/master/datasets/jfk_passengers>

**Data.** The Port Authority collects monthly data on domestic and
international passengers, cargo, flights, and aircraft equipment types
from each carrier operating at PANYNJ airports.

**Task.** This dataset can be analyzed using a change-point model with
different means before and after the change point, possibly including a
linear drift after the change point. A basic model is
$$ y_t = \mu_t + \epsilon_t, $$ where\*
$$ \epsilon_t \sim {N}(0,\sigma_t^2), $$ and
$$ \mu_t = \mu_1,\quad \sigma_t = \sigma_1 \qquad \text{if } t \leq \tau, $$
while
$$ \mu_t = \mu_2,\quad \sigma_t = \sigma_2 \qquad \text{if } t > \tau. $$
Here $\tau$ is an unknown change-point parameter.

More complex change-point models may also be considered and discussed.
Time series models are encouraged; for example, one may include
different trends and different variances before and after the change
point.

```{r}
#brief dataset inspection

rm(list = ls())
JFK <- read.csv("JFK.csv")
head(JFK)
tail(JFK)
dim(JFK)
JFK$time__raw <- as.Date(paste(JFK$time__raw, "-01", sep = ""))
plot(JFK$time__raw, JFK$series__raw, xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[1:36], JFK$series__raw[1:36], xlab = "time", ylab = "pass", main = "First 3 years of data")
plot(JFK$time__raw[408:468], JFK$series__raw[408:468], xlab = "time", ylab = "pass", main = "last 5 years of data")

# we try to average over the year to see the trend without the seasonal oscillation

n_years <- length(JFK$time__raw) / 12

years <- 1977:2015

# avraging over the months

annual_mean <- tapply(
  JFK$series__raw,
  rep(years, each = 12),
  mean
)

# plot

plot(
  1977:2015,
  annual_mean,
  type = "b",
  xlab = "year",
  ylab = "average passengers",
  main = "average annual passenger JFK"
)
```

```         
```

## Simple Changepoint Model

$$ y_t = \mu_t + \epsilon $$

```{r}
length(JFK$time__raw)
```

```{r}
library(rjags)
library(bayesplot)
library(jagsUI)
library(loo)
library(coda)
```

```{r}
modelChangePoint <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv)
    mu[i] <- m[J[i]]
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv))  # this is for computing the WAIC, see later
    #predictive
    xpred[i] ~ dnorm(mu[i], sinv)
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta
  m[1] ~ dnorm(2500000,0.0000000000001)
  m[2] ~ dnorm(4000000,0.0000000000001)
  sinv ~  dunif(0, 10)

}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw))

  inits <- list(
    list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
    list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8),
    list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 15),
    list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 16)
)

sampCP_mcmc <- jags(
  model.file = textConnection(modelChangePoint), data = myData,
  parameters.to.save = c("mu", "xpred", "Tau", "J", "LogLik", "sinv", "m"),
  n.adapt = 5000, n.iter = 25000, inits=inits, n.chains = 4, n.burnin = 5000
)


ssCP <- summary(sampCP_mcmc)
```

```{r}
# --- MCMC Diagnostics ---

# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc$Rhat[c("Tau", "sinv", "m")]
sampCP_mcmc$n.eff[c("Tau", "sinv", "m")]

# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc, "Tau")
plot(sampCP_mcmc, "sinv")
plot(sampCP_mcmc, "m")


# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc$samples[[1]][, c("Tau", "sinv", "m[1]", "m[2]")]), lag.max = 40)



```

```{r}
q1 <- ssCP["Tau", c("2.5%")]
q2 <- ssCP["Tau", c("97.5%")]
tmean <- ssCP["Tau", c("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc, "mu[329]")

```

```{r}
sampCP_mcmc$DIC
loglik_sampCP <- sampCP_mcmc$sims.list$LogLik # model with change point

waic_CP <- waic(loglik_sampCP)

waic_CP
```

## Switching variance

$$ y_t = \mu_t + \epsilon_t $$

```{r}
modelChangePoint_2v <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- m[J[i]]
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    #predictive
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta
  m[1] ~ dnorm(2.5e6,1e-13)
  m[2] ~ dnorm(4e6,1e-13)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)
  
  #to avoid numerical problem when calculating ESS
  sigma[1] <- 1/sqrt(sinv_p[1])
  sigma[2] <- 1/sqrt(sinv_p[2])


}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 23),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 42),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8)
)

sampCP_mcmc_2v <- jags(

  model.file = textConnection(modelChangePoint_2v), data = myData,
  parameters.to.save = c("mu", "xpred", "Tau", "J", "LogLik", "sigma", "m"),
  n.adapt = 5000, n.iter = 10000, inits=inits, n.chains = 4, n.burnin = 500
)


ssCP_2v <- summary(sampCP_mcmc_2v)
```

```{r}
# --- MCMC Diagnostics ---

# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc_2v$Rhat[c("Tau", "m")]
sampCP_mcmc_2v$Rhat$sigma

sampCP_mcmc_2v$n.eff[c("Tau", "m")]
sampCP_mcmc_2v$n.eff$sigma


# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc_2v, "Tau")
plot(sampCP_mcmc_2v, "sigma")
plot(sampCP_mcmc_2v, "m")



# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc_2v$samples[[1]][, c("Tau", "m[1]", "m[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
```

```{r}
q1 <- ssCP_2v["Tau", c("2.5%")]
q2 <- ssCP_2v["Tau", c("97.5%")]
tmean <- ssCP_2v["Tau", c("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_2v[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_2v[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_2v[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_2v[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_2v[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc_2v, "mu[329]")

```

```{r}
sampCP_mcmc_2v$DIC
loglik_sampCP_2v <- sampCP_mcmc_2v$sims.list$LogLik # flat model with change point and 2 variances

waic_CP_2v <- waic(loglik_sampCP_2v)

waic_CP_2v
```

## Linear Drift

$$ y_t = \mu_t + \epsilon_t $$ where $$ \mu_t = \alpha_t + \beta_t X $$

```{r}
modelChangePoint_lin <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- alpha[J[i]]+beta[J[i]]*i
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta

  alpha[1] ~ dnorm(0,0.000000000001)
  alpha[2] ~ dnorm(0,0.000000000001)
  beta[1] ~ dnorm(0,0.0000000000001)
  beta[2] ~ dnorm(0,0.0000000000001)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)
  
  #to avoid numerical problem when calculating ESS
  sigma[1] <- 1/sqrt(sinv_p[1])
  sigma[2] <- 1/sqrt(sinv_p[2])

}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 23),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 42),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8)
)

sampCP_mcmc_lin <- jags(

  model.file = textConnection(modelChangePoint_lin), data = myData,
  parameters.to.save = c("mu", "xpred", "alpha", "beta", "Tau", "J", "LogLik", "sigma"),
  n.adapt = 1000, n.iter = 10000,inits=inits, n.chains = 4, n.burnin = 2000
)


ssCP_lin <- summary(sampCP_mcmc_lin)
```

```{r}

# --- MCMC Diagnostics ---
# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc_lin$Rhat$Tau
sampCP_mcmc_lin$Rhat$alpha
sampCP_mcmc_lin$Rhat$beta
sampCP_mcmc_lin$Rhat$sigma
sampCP_mcmc_lin$n.eff$Tau
sampCP_mcmc_lin$n.eff$alpha
sampCP_mcmc_lin$n.eff$beta
sampCP_mcmc_lin$n.eff$sigma
# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc_lin, "Tau")
plot(sampCP_mcmc_lin, "alpha")
plot(sampCP_mcmc_lin, "beta")
plot(sampCP_mcmc_lin, "sigma")
# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc_lin$samples[[1]][, c("Tau", "alpha[1]", "alpha[2]", "beta[1]", "beta[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
```

```{r}
q1 <- ssCP_lin["Tau", c("2.5%")]
q2 <- ssCP_lin["Tau", c("97.5%")]
tmean <- ssCP_lin["Tau", c("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_lin[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_lin[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_lin[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_lin[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_lin[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc_lin, "mu[329]")

```

```{r}
sampCP_mcmc_lin$DIC
loglik_lin <- sampCP_mcmc_lin$sims.list$LogLik # model with change point

waic_lin <- waic(loglik_lin)

waic_lin
```

## Periodic model

$$ y_t = \mu_t + \epsilon_t $$ where $$ \mu_t = \alpha_t + \beta_t X $$
and
$$  \text{where } \epsilon_t \sim {N}(\gamma \cos(\frac{2\pi m}{12}) + \delta \sin(\frac{2\pi m}{12}),\sigma_t^2) $$

```{r}
modelChangePoint_period <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- alpha[J[i]]+beta[J[i]]*i + gamma[J[i]]*cos(w*i) + delta[J[i]]*sin(w*i)
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }

  pi <- 3.14159265358979

  w <- 2 * pi / 12
  
  # Prior for eps[i]
  gamma[1] ~  dnorm(0,0.000000000001)
  delta[1] ~  dnorm(0,0.000000000001)
  gamma[2] ~  dnorm(0,0.000000000001)
  delta[2] ~  dnorm(0,0.000000000001)

  # Prior for d[i]
  Tau ~ dunif(1, n+10)
  
  # Prior for beta
  alpha[1] ~ dnorm(0,0.000000000001)
  alpha[2] ~ dnorm(0,0.000000000001)
  beta[1] ~ dnorm(0,0.0000000000001)
  beta[2] ~ dnorm(0,0.0000000000001)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)
  
  #to avoid numerical problem when calculating ESS
  sigma[1] <- 1/sqrt(sinv_p[1])
  sigma[2] <- 1/sqrt(sinv_p[2])
  
}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 15),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 16)
)

sampCP_mcmc_period <- jags(
  model.file = textConnection(modelChangePoint_period), data = myData,
  parameters.to.save = c("mu", "xpred", "alpha", "beta", "sigma", "gamma", "delta", "Tau", "J", "LogLik", "sinv_p"),
  n.adapt = 1000, n.iter = 20000, inits=inits, n.chains = 4, n.burnin = 4000
)


ssCP_per <- summary(sampCP_mcmc_period)
```

```{r}

# --- MCMC Diagnostics ---
# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc_period$Rhat$Tau
sampCP_mcmc_period$Rhat$alpha
sampCP_mcmc_period$Rhat$beta
sampCP_mcmc_period$Rhat$sigma
sampCP_mcmc_period$Rhat$gamma
sampCP_mcmc_period$Rhat$delta

sampCP_mcmc_period$n.eff$Tau
sampCP_mcmc_period$n.eff$alpha
sampCP_mcmc_period$n.eff$beta
sampCP_mcmc_period$n.eff$sigma
sampCP_mcmc_period$n.eff$gamma
sampCP_mcmc_period$n.eff$delta



# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc_period, "Tau")
plot(sampCP_mcmc_period, "alpha")
plot(sampCP_mcmc_period, "beta")
plot(sampCP_mcmc_period, "sigma")
plot(sampCP_mcmc_period, "gamma")
plot(sampCP_mcmc_period, "delta")

# Posterior mean estimates of seasonal amplitude before and after change point
gamma1 <- ssCP_per["gamma[1]", "mean"]
gamma2 <- ssCP_per["gamma[2]", "mean"]
delta1 <- ssCP_per["delta[1]", "mean"]
delta2 <- ssCP_per["delta[2]", "mean"]

A1 <- sqrt(gamma1^2 + delta1^2)
A2 <- sqrt(gamma2^2 + delta2^2)

print(paste("Seasonal amplitude before change point:", round(A2, 2)))
print(paste("Seasonal amplitude after change point:", round(A1, 2)))


# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc_period$samples[[1]][, c( "alpha[1]", "alpha[2]", "beta[1]", "beta[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
autocorr.plot(as.mcmc(sampCP_mcmc_period$samples[[1]][, c( "Tau", "gamma[1]", "gamma[2]" ,"delta[1]", "delta[2]")]), lag.max = 40)
```

```{r}
q1 <- ssCP_per["Tau", c("2.5%")]
q2 <- ssCP_per["Tau", c("97.5%")]
tmean <- ssCP_per["Tau", c("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, xlim=c(200,250), ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_per[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_per[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_per[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_per[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_per[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc_period, "mu[329]")

```

```{r}
sampCP_mcmc_period$DIC
loglik_per <- sampCP_mcmc_period$sims.list$LogLik # model with change point

waic_per <- waic(loglik_per)

waic_per
```

```{r}
modelAR <-"model {
  ## parameters: alpha,tau,m0
  #likeliohood 
  mu[1]<-Y[1]
  Yp[1]=mu[1]
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i],tau)
    mu[i]<-m0+alpha*Y[i-1]
    Yp[i] ~ dnorm(mu[i],tau) # prediction in sample
  }
  # prediction out of sample 
  ypOut[1] ~dnorm(m0+alpha*Y[N],tau) 
  for(k in 2:Npred){
  ypOut[k] ~dnorm(m0+alpha*ypOut[k-1],tau) 
  }
  sigma2<-1/tau
  #prior 
  alpha ~ dunif(-1.5,1.5)
  tau ~ dgamma(0.1, 10)
  m0 ~dnorm(0.0, 1.0E-10)
}"
```

```{r}
myData <- list(Y = JFK$series__raw, 
               N = length(JFK$series__raw),
               Npred = 12)  # quanti passi vuoi prevedere nel futuro

sampAR_mcmc <- jags(
  model.file = textConnection(modelAR), 
  data = myData,
  parameters.to.save = c("mu", "Yp", "ypOut", "alpha", "m0", "tau", "sigma2"),
  n.adapt = 1000, 
  n.iter = 5000, 
  n.chains = 2, 
  n.burnin = 2000
)

ssAR <- summary(sampAR_mcmc)
```

```{r}
#q1 <- ssAR["m0", c("2.5%")]
#q2 <- ssAR["m0", c("97.5%")]
#mu_mean <- ssAR["", muc("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, xlim = c(200, 250), ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), ssAR[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssAR[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssAR[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssAR[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssAR[1:468, "mean"], col = "magenta", lwd = 1.0)
#abline(v = tmean, col = "red", lwd = 1.5)
#abline(v = q1, col = "red", lwd = 1.5, lty = 3)
#abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampAR_mcmc, "m0")
plot(sampAR_mcmc, "alpha")
#plot(sampCP_mcmc_period, "mu[293]")
#plot(sampCP_mcmc_period, "xpred[1]")
#plot(sampCP_mcmc_period, "alpha[1]")
#plot(sampCP_mcmc_period, "alpha[2]")
#plot(sampCP_mcmc_period, "beta[1]")
#plot(sampCP_mcmc_period, "beta[2]")
```

```{r}
modelAR_period <- "model{

##################################
# CHANGE POINT
##################################

for(t in 1:n){
d[t] <- step(Tau - t)
}

##################################
# FIRST OBSERVATION
##################################

J[1] <- d[1] + 1

mu[1] <-
alpha[J[1]]
+ beta[J[1]] * 1
+ gamma * cos(w * 1)
+ delta * sin(w * 1)

sinv[1] <- sinv_p[J[1]]

x[1] ~ dnorm(mu[1], sinv[1])

LogLik[1] <- log(dnorm(x[1],mu[1],sinv[1]))

xpred[1] ~ dnorm(mu[1],sinv[1])

##################################
# AR(1)
##################################
for(i in 2:n){

J[i] <- d[i] + 1

sinv[i] <- sinv_p[J[i]]

mu[i] <-
alpha[J[i]]
+ beta[J[i]] * i
+ gamma * cos(w * i)
+ delta * sin(w * i)

meanAR[i] <-
mu[i]
+ a * (x[i-1] - mu[i-1])

x[i] ~ dnorm(meanAR[i], sinv[i])

LogLik[i] <- log(dnorm(x[i],meanAR[i],sinv[i]))

xpred[i] ~ dnorm(meanAR[i],sinv[i])
}

pi <- 3.14159265358979
w <- 2*pi/12
##################################
# PRIORS
##################################
gamma ~ dnorm(0,1.0E-12)
delta ~ dnorm(0,1.0E-12)
alpha[1] ~ dnorm(0,1.0E-12)
alpha[2] ~ dnorm(0,1.0E-12)
beta[1] ~ dnorm(0,1.0E-13)
beta[2] ~ dnorm(0,1.0E-13)
Tau ~ dunif(1, n+10)
a ~ dunif(-1,1)
sinv_p[1] ~ dunif(0,10)
sinv_p[2] ~ dunif(0,10)

#to avoid numerical problem when calculating ESS
sigma[1] <- 1/sqrt(sinv_p[1])
sigma[2] <- 1/sqrt(sinv_p[2])


}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #


inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 15),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 16)
)


sampCP_mcmc_AR_period <- jags(
model.file = textConnection(modelAR_period), data = myData,
parameters.to.save = c("mu", "xpred", "alpha", "beta", "gamma", "delta", "Tau", "a", "J", "LogLik", "sigma"),
n.adapt = 1000, n.iter = 10000, inits=inits, n.chains = 4, n.burnin = 2000
)
ssCP_AR_period <- summary(sampCP_mcmc_AR_period)
```

```{r}
# --- MCMC Diagnostics ---
# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc_AR_period$Rhat$Tau
sampCP_mcmc_AR_period$Rhat$alpha
sampCP_mcmc_AR_period$Rhat$beta
sampCP_mcmc_AR_period$Rhat$gamma
sampCP_mcmc_AR_period$Rhat$delta
sampCP_mcmc_AR_period$Rhat$a
sampCP_mcmc_AR_period$Rhat$sigma
sampCP_mcmc_AR_period$n.eff$Tau
sampCP_mcmc_AR_period$n.eff$alpha
sampCP_mcmc_AR_period$n.eff$beta
sampCP_mcmc_AR_period$n.eff$gamma
sampCP_mcmc_AR_period$n.eff$delta
sampCP_mcmc_AR_period$n.eff$a
sampCP_mcmc_AR_period$n.eff$sigma

# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc_AR_period, "Tau")
plot(sampCP_mcmc_AR_period, "alpha")
plot(sampCP_mcmc_AR_period, "beta")
plot(sampCP_mcmc_AR_period, "gamma")
plot(sampCP_mcmc_AR_period, "delta")
plot(sampCP_mcmc_AR_period, "a")
plot(sampCP_mcmc_AR_period, "sigma")

# Posterior mean estimate of seasonal amplitude (single gamma/delta, not split by regime)
gamma_m <- ssCP_AR_period["gamma", "mean"]
delta_m <- ssCP_AR_period["delta", "mean"]
A <- sqrt(gamma_m^2 + delta_m^2)
print(paste("Seasonal amplitude:", round(A, 2)))

# Posterior mean of AR(1) coefficient
print(paste("AR(1) coefficient (a):", round(ssCP_AR_period["a", "mean"], 3)))

# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc_AR_period$samples[[1]][, c("alpha[1]", "alpha[2]", "beta[1]", "beta[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
autocorr.plot(as.mcmc(sampCP_mcmc_AR_period$samples[[1]][, c("Tau", "gamma", "delta", "a")]), lag.max = 40)
```

```{r}
q1 <- ssCP_AR_period["Tau", c("2.5%")]
q2 <- ssCP_AR_period["Tau", c("97.5%")]
tmean <- ssCP_AR_period["Tau", c("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, xlim = c(300, 350), ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_AR_period[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_AR_period[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_AR_period[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_AR_period[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_AR_period[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc_AR_period, "mu[329]")
```

```         
```

```{r}
sampCP_mcmc_AR_period$DIC
loglik_ar_period <- sampCP_mcmc_AR_period$sims.list$LogLik # model with change point
waic_ar_period <- waic(loglik_ar_period)
waic_ar_period
```

### ARX(1)

```{r}
modelARX_period <- "model{

for(t in 1:n){
d[t] <- step(Tau - t)
}

J[1] <- d[1] + 1

mu[1] <- alpha[J[1]] + beta[J[1]] * 1

err[1] <- gamma * cos(w * 1) + delta * sin(w * 1)

sinv[1] <- sinv_p[J[1]]

meanARX[1] <- mu[1] + err[1]

x[1] ~ dnorm(meanARX[1], sinv[1])

LogLik[1] <- log(dnorm(x[1],meanARX[1],sinv[1]))

xpred[1] ~ dnorm(meanARX[1],sinv[1])

# AR(1)

for(i in 2:n){

J[i] <- d[i] + 1

sinv[i] <- sinv_p[J[i]]

mu[i] <- alpha[J[i]] + beta[J[i]] * i

err[i] <- 
gamma * cos(w * i)
+ delta * sin(w * i)
+ a * (x[i-1] - mu[i-1])

meanARX[i] <-
mu[i]
+ err[i]

x[i] ~ dnorm(meanARX[i], sinv[i])

LogLik[i] <- log(dnorm(x[i],meanARX[i],sinv[i]))

xpred[i] ~ dnorm(meanARX[i],sinv[i])
}

pi <- 3.14159265358979
w <- 2*pi/12

# PRIORS

gamma ~ dnorm(0,1.0E-12)
delta ~ dnorm(0,1.0E-12)
alpha[1] ~ dnorm(0,1.0E-12)
alpha[2] ~ dnorm(0,1.0E-12)
beta[1] ~ dnorm(0,1.0E-13)
beta[2] ~ dnorm(0,1.0E-13)
Tau ~ dnorm(300, 0.1)
a ~ dunif(-10,10)
sinv_p[1] ~ dunif(0,10)
sinv_p[2] ~ dunif(0,10)

#to avoid numerical problem when calculating ESS
sigma[1] <- 1/sqrt(sinv_p[1])
sigma[2] <- 1/sqrt(sinv_p[2])


}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #


inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 23),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 42),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8)
)

sampCP_mcmc_ARX_period <- jags(
model.file = textConnection(modelARX_period), data = myData,
parameters.to.save = c("mu", "meanARX", "xpred", "alpha", "beta", "gamma", "delta", "Tau", "a", "J", "LogLik", "sigma"),
n.adapt = 1000, n.iter = 25000, inits=inits, n.chains = 4, n.burnin = 5000
)
ssCP_ARX <- summary(sampCP_mcmc_ARX_period)
```

```{r}
# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_mcmc_ARX_period$Rhat$Tau
sampCP_mcmc_ARX_period$Rhat$alpha
sampCP_mcmc_ARX_period$Rhat$beta
sampCP_mcmc_ARX_period$Rhat$gamma
sampCP_mcmc_ARX_period$Rhat$delta
sampCP_mcmc_ARX_period$Rhat$a
sampCP_mcmc_ARX_period$Rhat$sigma
sampCP_mcmc_ARX_period$n.eff$Tau
sampCP_mcmc_ARX_period$n.eff$alpha
sampCP_mcmc_ARX_period$n.eff$beta
sampCP_mcmc_ARX_period$n.eff$gamma
sampCP_mcmc_ARX_period$n.eff$delta
sampCP_mcmc_ARX_period$n.eff$a
sampCP_mcmc_ARX_period$n.eff$sigma

# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_mcmc_ARX_period, "Tau")
plot(sampCP_mcmc_ARX_period, "alpha")
plot(sampCP_mcmc_ARX_period, "beta")
plot(sampCP_mcmc_ARX_period, "gamma")
plot(sampCP_mcmc_ARX_period, "delta")
plot(sampCP_mcmc_ARX_period, "a")
plot(sampCP_mcmc_ARX_period, "sigma")

# Posterior mean estimate of seasonal amplitude (single gamma/delta, not split by regime)
gamma_m <- ssCP_ARX["gamma", "mean"]
delta_m <- ssCP_ARX["delta", "mean"]
A <- sqrt(gamma_m^2 + delta_m^2)
print(paste("Seasonal amplitude:", round(A, 2)))

# Posterior mean of AR(1) coefficient
print(paste("AR(1) coefficient (a):", round(ssCP_ARX["a", "mean"], 3)))

# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_mcmc_ARX_period$samples[[1]][, c("alpha[1]", "alpha[2]", "beta[1]", "beta[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
autocorr.plot(as.mcmc(sampCP_mcmc_ARX_period$samples[[1]][, c("Tau", "gamma", "delta", "a")]), lag.max = 40)
```

```{r}
q1 <- ssCP_ARX["Tau", c("2.5%")]
q2 <- ssCP_ARX["Tau", c("97.5%")]
tmean <- ssCP_ARX["Tau", c("mean")]

mu_rows    <- grep("^meanARX\\[", rownames(ssCP_ARX))
xpred_rows <- grep("^xpred\\[", rownames(ssCP_ARX))

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_ARX[mu_rows, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_ARX[mu_rows, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_ARX[xpred_rows, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_ARX[xpred_rows, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_ARX[mu_rows, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)
plot(sampCP_mcmc_ARX_period, "meanARX[297]")

```

```{r}
sampCP_mcmc_ARX_period$DIC
loglik_ARX <- sampCP_mcmc_ARX_period$sims.list$LogLik # model with change point
waic_ARX <- waic(loglik_ARX)
waic_ARX
```

### Model Selection

```{r}
library(loo)

# Compute LOO-CV for each model
loo_sampCP <- loo(loglik_sampCP)
loo_sampCP_2v <- loo(loglik_sampCP_2v)
loo_lin <- loo(loglik_lin)
loo_per <- loo(loglik_per)
loo_ar_period <- loo(loglik_ar_period)
loo_ARX <- loo(loglik_ARX)

# Compute WAIC for each model
waic_sampCP    <- waic(loglik_sampCP)
waic_sampCP_2v <- waic(loglik_sampCP_2v)
waic_lin       <- waic(loglik_lin)
waic_per       <- waic(loglik_per)
waic_ar_period <- waic(loglik_ar_period)
waic_ARX       <- waic(loglik_ARX)
```

```{r}
# Compare the models
loo_comparison <- loo_compare(list(
  SampCP = loo_sampCP,
  SampCP2v = loo_sampCP_2v,
  Lin = loo_lin,
  Periodic = loo_per,
  AR = loo_ar_period,
  ARX = loo_ARX
))

waic_comparison <- loo_compare(list(
  SampCP   = waic_sampCP,
  SampCP2v = waic_sampCP_2v,
  Lin      = waic_lin,
  Periodic = waic_per,
  AR       = waic_ar_period,
  ARX      = waic_ARX
))

model_names <- c("SampCP", "SampCP2v", "Lin", "Periodic", "AR", "ARX")

loo_list  <- list(loo_sampCP, loo_sampCP_2v, loo_lin, loo_per, loo_ar_period, loo_ARX)
waic_list <- list(waic_sampCP, waic_sampCP_2v, waic_lin, waic_per, waic_ar_period, waic_ARX)

ic_table <- data.frame(
  Model   = model_names,
  LOOIC   = sapply(loo_list,  \(x) round(x$estimates["looic", "Estimate"], 2)),
  SE_LOO  = sapply(loo_list,  \(x) round(x$estimates["looic", "SE"], 2)),
  WAIC    = sapply(waic_list, \(x) round(x$estimates["waic",  "Estimate"], 2)),
  SE_WAIC = sapply(waic_list, \(x) round(x$estimates["waic",  "SE"], 2))
)

print(ic_table[order(ic_table$LOOIC), ])
```

```{r}
modelAR_period_PRED <- "model{
#change point
for(t in 1:n_train){
  d[t] <- step(Tau - t)
}
#first observation
J[1] <- d[1] + 1
mu[1] <- alpha[J[1]] + beta[J[1]] * 1 + gamma * cos(w * 1) + delta * sin(w * 1)
sinv[1] <- sinv_p[J[1]]
x[1] ~ dnorm(mu[1], sinv[1])
LogLik[1] <- log(dnorm(x[1], mu[1], sinv[1]))
xpred[1] ~ dnorm(mu[1], sinv[1])

#likelihood
for(i in 2:n_train){
  J[i] <- d[i] + 1
  sinv[i] <- sinv_p[J[i]]
  mu[i] <- alpha[J[i]] + beta[J[i]] * i + gamma * cos(w * i) + delta * sin(w * i)
  meanAR[i] <- mu[i] + a * (x[i-1] - mu[i-1])
  x[i] ~ dnorm(meanAR[i], sinv[i])
  LogLik[i] <- log(dnorm(x[i], meanAR[i], sinv[i]))
  xpred[i] ~ dnorm(meanAR[i], sinv[i])
}

#out-of-sample prediction ( we assume t > Tau)
mupred[1] <- alpha[1] + beta[1] * (n_train + 1) + gamma * cos(w * (n_train + 1)) + delta * sin(w * (n_train + 1))
ypOut_mean[1] <- mupred[1] + a * (x[n_train] - mu[n_train])
ypOut[1] ~ dnorm(ypOut_mean[1], sinv_p[1])

for(k in 2:Npred){
  mupred[k] <- alpha[1] + beta[1] * (n_train + k) + gamma * cos(w * (n_train + k)) + delta * sin(w * (n_train + k))
  ypOut_mean[k] <- mupred[k] + a * (ypOut[k-1] - mupred[k-1])
  ypOut[k] ~ dnorm(ypOut_mean[k], sinv_p[1])
}

pi <- 3.14159265358979
w <- 2*pi/12

#priors
gamma ~ dnorm(0,1.0E-12)
delta ~ dnorm(0,1.0E-12)
alpha[1] ~ dnorm(0,1.0E-12)
alpha[2] ~ dnorm(0,1.0E-12)
beta[1] ~ dnorm(0,1.0E-13)
beta[2] ~ dnorm(0,1.0E-13)
Tau ~ dnorm(300, 0.1)
a ~ dunif(-1,1)
sinv_p[1] ~ dunif(0,10)
sinv_p[2] ~ dunif(0,10)

#to avoid numerical problem when calculating ESS
sigma[1] <- 1/sqrt(sinv_p[1])
sigma[2] <- 1/sqrt(sinv_p[2])
}"
  
```

```{r}
n_train <- 432  #we split defore the last year
Npred <- 36

inits <- list(
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 4),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 8),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 15),
  list(.RNG.name = "base::Wichmann-Hill",  .RNG.seed = 16)
)


myData_train <- list(x = JFK$series__raw[1:n_train], n_train = n_train, Npred = Npred)

sampCP_train <- jags(
  model.file = textConnection(modelAR_period_PRED), data = myData_train,
  parameters.to.save = c("mu", "xpred", "alpha", "beta", "gamma", "delta", "Tau", "a", "sigma", "ypOut"),
  n.adapt = 1000, n.iter = 10000, inits = inits, n.chains = 4, n.burnin = 2000
)
```

```{r}
# --- MCMC Diagnostics ---
# R-hat and ESS for key parameters (already computed by jagsUI)
sampCP_train$Rhat$Tau
sampCP_train$Rhat$alpha
sampCP_train$Rhat$beta
sampCP_train$Rhat$gamma
sampCP_train$Rhat$delta
sampCP_train$Rhat$a
sampCP_train$Rhat$sigma
sampCP_train$Rhat$ypOut
sampCP_train$n.eff$Tau
sampCP_train$n.eff$alpha
sampCP_train$n.eff$beta
sampCP_train$n.eff$gamma
sampCP_train$n.eff$delta
sampCP_train$n.eff$a
sampCP_train$n.eff$sigma
sampCP_train$n.eff$ypOut

# Trace plot + posterior density (R-hat shown automatically)
plot(sampCP_train, "Tau")
plot(sampCP_train, "alpha")
plot(sampCP_train, "beta")
plot(sampCP_train, "gamma")
plot(sampCP_train, "delta")
plot(sampCP_train, "a")
plot(sampCP_train, "sigma")

# Posterior mean estimate of seasonal amplitude (single gamma/delta, not split by regime)
ssCP_train <- summary(sampCP_train)
gamma_m <- ssCP_train["gamma", "mean"]
delta_m <- ssCP_train["delta", "mean"]
A <- sqrt(gamma_m^2 + delta_m^2)
print(paste("Seasonal amplitude:", round(A, 2)))

# Posterior mean of AR(1) coefficient
print(paste("AR(1) coefficient (a):", round(ssCP_train["a", "mean"], 3)))

# Autocorrelation for key parameters
autocorr.plot(as.mcmc(sampCP_train$samples[[1]][, c("alpha[1]", "alpha[2]", "beta[1]", "beta[2]", "sigma[1]", "sigma[2]")]), lag.max = 40)
autocorr.plot(as.mcmc(sampCP_train$samples[[1]][, c("Tau", "gamma", "delta", "a")]), lag.max = 40)
```

```{r}
# In-sample plot (fit on training data)
q1 <- ssCP_train["Tau", c("2.5%")]
q2 <- ssCP_train["Tau", c("97.5%")]
tmean <- ssCP_train["Tau", c("mean")]

mu_rows    <- grep("^mu\\[", rownames(ssCP_train))
xpred_rows <- grep("^xpred\\[", rownames(ssCP_train))

plot(1:n_train, JFK$series__raw[1:n_train], col = "orange", type = "p", lwd = 2.0,
     #xlim = c(0, 50),
     ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5),
     xlab = "time", ylab = "Y")
lines(1:n_train, JFK$series__raw[1:n_train], type = "l", col = "orange")
lines(1:n_train, ssCP_train[mu_rows, "2.5%"], type = "l", lty = 2, col = "blue")
lines(1:n_train, ssCP_train[mu_rows, "97.5%"], type = "l", lty = 2, col = "blue")
lines(1:n_train, ssCP_train[xpred_rows, "2.5%"], type = "l", lty = 2, col = "green")
lines(1:n_train, ssCP_train[xpred_rows, "97.5%"], type = "l", lty = 2, col = "green")
points(1:n_train, ssCP_train[mu_rows, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c("Data", "Model", "Credible interval", "Predictive Interval", "Changepoint")
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# Out-of-sample prediction plot
ypOut_rows <- grep("^ypOut\\[", rownames(ssCP_train))

yp_pred <- ssCP_train[ypOut_rows, "mean"]
q1_pred <- ssCP_train[ypOut_rows, "2.5%"]
q2_pred <- ssCP_train[ypOut_rows, "97.5%"]

n_total <- n_train + Npred
future_t <- (n_train + 1):n_total
y_true <- JFK$series__raw[future_t]

plot(1:n_total, JFK$series__raw[1:n_total], col = "red", type = "l",
     ylim = c(min(q1_pred, JFK$series__raw[1:n_total]) - 0.5,
              max(q2_pred, JFK$series__raw[1:n_total]) + 0.5),
     xlim = c(n_total-48, n_total),
     xlab = "time", ylab = "Y",
     main = "Out-of-sample prediction")
points(future_t, JFK$series__raw[future_t], col = "orange", pch = 19)  # true test values
abline(v = n_train, col = "orange", lwd = 1.5)
points(future_t, yp_pred, pch = "*", col = "magenta", cex = 1.5)
lines(future_t, q1_pred, col = "magenta", lwd = 1.5, lty = 2)
lines(future_t, q2_pred, col = "magenta", lwd = 1.5, lty = 2)

labels2 <- c("Observed data", "True test values", "Predicted mean", "95% Predictive interval")
legend("topleft", legend = labels2, col = c("red", "orange", "magenta", "magenta"),
       lty = c(1, NA, NA, 2), pch = c(NA, 19, 8, NA), cex = 0.75)

length(y_true)
length(q1_pred)
length(q2_pred)

inside <- (y_true >= q1_pred) & (y_true <= q2_pred)


coverage <- mean(inside)

coverage
```
