---
title: "Final Project"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

# Final Project: JFK Passengers

**Source.**
<https://github.com/alan-turing-institute/TCPD/tree/master/datasets/jfk_passengers>

**Data.** The Port Authority collects monthly data on domestic and
international passengers, cargo, flights, and aircraft equipment types
from each carrier operating at PANYNJ airports.

**Task.** This dataset can be analyzed using a change-point model with
different means before and after the change point, possibly including a
linear drift after the change point. A basic model is
$$ y_t = \mu_t + \epsilon_t, $$ where\*
$$ \epsilon_t \sim {N}(0,\sigma_t^2), $$ and
$$ \mu_t = \mu_1,\quad \sigma_t = \sigma_1 \qquad \text{if } t \leq \tau, $$
while
$$ \mu_t = \mu_2,\quad \sigma_t = \sigma_2 \qquad \text{if } t > \tau. $$
Here $\tau$ is an unknown change-point parameter.

More complex change-point models may also be considered and discussed.
Time series models are encouraged; for example, one may include
different trends and different variances before and after the change
point.

```{r}
rm(list = ls())
JFK <- read.csv("JFK.csv")
head(JFK)
tail(JFK)
dim(JFK)
JFK$time__raw <- as.Date(paste(JFK$time__raw, "-01", sep = ""))
plot(JFK$time__raw, JFK$series__raw, xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[1:13], JFK$series__raw[1:13], xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[1:36], JFK$series__raw[1:36], xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[400:468], JFK$series__raw[400:468], xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[300:468], JFK$series__raw[300:468], xlab = "time", ylab = "pass", main = "JFK")
plot(JFK$time__raw[275:400], JFK$series__raw[275:400], xlab = "time", ylab = "pass", main = "JFK")

# we try to average over the year to see the trend without the seasonal oscillation

n_years <- length(JFK$time__raw) / 12

years <- 1977:2015

# avraging over the months

annual_mean <- tapply(
  JFK$series__raw,
  rep(years, each = 12),
  mean
)

# plot

plot(
  1977:2015,
  annual_mean,
  type = "b",
  xlab = "year",
  ylab = "average passengers",
  main = "average annual passenger JFK"
)
```

## Simple Changepoint Model

$$ y_t = \mu_t + \epsilon $$

```{r}
mean(JFK$series__raw[1:350])
mean(JFK$series__raw[350:468])
var(JFK$series__raw)
length(JFK)
```

```{r}
library(rjags)
library(bayesplot)
library(jagsUI)
library(loo)
```

```{r}
modelChangePoint <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv)
    mu[i] <- m[J[i]]
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv))  # this is for computing the WAIC, see later
    #predictive
    xpred[i] ~ dnorm(mu[i], sinv)
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta
  m[1] ~ dnorm(2500000,0.0000000000001)
  m[2] ~ dnorm(4000000,0.0000000000001)
  sinv ~  dunif(0,10)

}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

sampCP_mcmc <- jags(
  model.file = textConnection(modelChangePoint), data = myData,
  parameters.to.save = c("mu", "xpred", "Tau", "J", "LogLik", "sinv"),
  n.adapt = 1000, n.iter = 10000, n.chains = 2, n.burnin = 2000
)


ssCP <- summary(sampCP_mcmc)
```

```{r}
q1 <- ssCP["Tau", c("2.5%")]
q2 <- ssCP["Tau", c("97.5%")]
tmean <- ssCP["Tau", c("mean")]
mtrueT <- c(rep(2354015, 330), rep(4070000, 168))

plot(mtrueT[1:468], col = "black", type = "l", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "p", col = "orange")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampCP_mcmc, "Tau")
plot(sampCP_mcmc, "mu[1]")
plot(sampCP_mcmc, "mu[329]")
plot(sampCP_mcmc, "xpred[1]")
```

```{r}
sampCP_mcmc$DIC
loglik <- sampCP_mcmc$sims.list$LogLik # model with change point

waic <- waic(loglik)

waic
```

## Switching variance

$$ y_t = \mu_t + \epsilon_t $$

```{r}
modelChangePoint_2v <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- m[J[i]]
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    #predictive
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta
  m[1] ~ dnorm(2500000,0.0000000000001)
  m[2] ~ dnorm(4000000,0.0000000000001)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)


}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

sampCP_mcmc_2v <- jags(
  model.file = textConnection(modelChangePoint_2v), data = myData,
  parameters.to.save = c("mu", "xpred", "Tau", "J", "LogLik", "sinv"),
  n.adapt = 1000, n.iter = 10000, n.chains = 2, n.burnin = 2000
)


ssCP_2v <- summary(sampCP_mcmc_2v)
```

```{r}
q1 <- ssCP_2v["Tau", c("2.5%")]
q2 <- ssCP_2v["Tau", c("97.5%")]
tmean <- ssCP_2v["Tau", c("mean")]
mtrueT <- c(rep(2354015, 330), rep(4070000, 168))

plot(mtrueT[1:468], col = "black", type = "l", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "p", col = "orange")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP_2v[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_2v[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP_2v[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP_2v[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP_2v[1:468, "mean"], col = "magenta", lwd = 0.75)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampCP_mcmc_2v, "Tau")
plot(sampCP_mcmc_2v, "mu[1]")
plot(sampCP_mcmc_2v, "mu[329]")
plot(sampCP_mcmc_2v, "xpred[1]")
```

```{r}
sampCP_mcmc_2v$DIC
loglik <- sampCP_mcmc_2v$sims.list$LogLik # flat model with change point and 2 variances

waic <- waic(loglik)

waic
```

## Linear Drift

$$ y_t = \mu_t + \epsilon_t $$ where $$ \mu_t = \alpha_t + \beta_t X $$

```{r}
modelChangePoint_lin <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- alpha[J[i]]+beta[J[i]]*i
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }


  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta

  alpha[1] ~ dnorm(0,0.000000000001)
  alpha[2] ~ dnorm(0,0.000000000001)
  beta[1] ~ dnorm(0,0.0000000000001)
  beta[2] ~ dnorm(0,0.0000000000001)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)

}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

sampCP_mcmc_lin <- jags(
  model.file = textConnection(modelChangePoint_lin), data = myData,
  parameters.to.save = c("mu", "xpred", "alpha", "beta", "Tau", "J", "LogLik", "sinv"),
  n.adapt = 1000, n.iter = 10000, n.chains = 2, n.burnin = 2000
)


ssCP <- summary(sampCP_mcmc_lin)
```

```{r}
q1 <- ssCP["Tau", c("2.5%")]
q2 <- ssCP["Tau", c("97.5%")]
tmean <- ssCP["Tau", c("mean")]
mtrueT <- c(rep(2354015, 330), rep(4070000, 168))

plot(mtrueT[1:468], col = "black", type = "l", lwd = 2.0, ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "p", col = "orange")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP[1:468, "mean"], col = "magenta", lwd = 1.0)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampCP_mcmc_lin, "Tau")
plot(sampCP_mcmc_lin, "mu[1]")
plot(sampCP_mcmc_lin, "mu[328]")
plot(sampCP_mcmc_lin, "xpred[1]")
plot(sampCP_mcmc_lin, "alpha[1]")
plot(sampCP_mcmc_lin, "alpha[2]")
plot(sampCP_mcmc_lin, "beta[1]")
plot(sampCP_mcmc_lin, "beta[2]")
```

```{r}
sampCP_mcmc_lin$DIC
loglik <- sampCP_mcmc_lin$sims.list$LogLik # model with change point

waic <- waic(loglik)

waic
```

## Periodic model

$$ y_t = \mu_t + \epsilon_t $$ where $$ \mu_t = \alpha_t + \beta_t X $$
and
$$  \text{where } \epsilon_t \sim {N}(\gamma \cos(\frac{2\pi m}{12}) + \delta \sin(\frac{2\pi m}{12}),\sigma_t^2) $$

```{r}
modelChangePoint_period <-
  "model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv[i])
    sinv[i] <- sinv_p[J[i]]
    mu[i] <- alpha[J[i]]+beta[J[i]]*i + gamma*cos(w*i) + delta*sin(w*i)
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv[i]))  # this is for computing the WAIC, see later
    xpred[i] ~ dnorm(mu[i], sinv[i])
  }

  pi <- 3.14159265358979

  w <- 2 * pi / 12
  
  # Prior for eps[i]
  gamma ~  dnorm(0,0.000000000001)
  delta ~  dnorm(0,0.000000000001)

  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  
  # Prior for beta
  alpha[1] ~ dnorm(0,0.000000000001)
  alpha[2] ~ dnorm(0,0.000000000001)
  beta[1] ~ dnorm(0,0.0000000000001)
  beta[2] ~ dnorm(0,0.0000000000001)
  sinv_p[1] ~  dunif(0,10)
  sinv_p[2] ~  dunif(0,10)
}"
```

```{r}
myData <- list(x = JFK$series__raw, n = length(JFK$series__raw)) #

sampCP_mcmc_period <- jags(
  model.file = textConnection(modelChangePoint_period), data = myData,
  parameters.to.save = c("mu", "xpred", "alpha", "beta", "gamma", "delta", "Tau", "J", "LogLik", "sinv"),
  n.adapt = 1000, n.iter = 10000, n.chains = 2, n.burnin = 2000
)


ssCP <- summary(sampCP_mcmc_period)
```

```{r}
q1 <- ssCP["Tau", c("2.5%")]
q2 <- ssCP["Tau", c("97.5%")]
tmean <- ssCP["Tau", c("mean")]
mtrueT <- c(rep(2354015, 330), rep(4070000, 168))

plot(mtrueT[1:468], col = "black", type = "l", lwd = 2.0,xlim=c(0, 10), ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), JFK$series__raw, type = "p", col = "orange")
lines(c(1:468), JFK$series__raw, type = "l", col = "orange")
lines(c(1:468), ssCP[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssCP[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssCP[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssCP[1:468, "mean"], col = "magenta", lwd = 1.0)
abline(v = tmean, col = "red", lwd = 1.5)
abline(v = q1, col = "red", lwd = 1.5, lty = 3)
abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampCP_mcmc_period, "Tau")
plot(sampCP_mcmc_period, "mu[1]")
plot(sampCP_mcmc_period, "mu[293]")
plot(sampCP_mcmc_period, "xpred[1]")
plot(sampCP_mcmc_period, "alpha[1]")
plot(sampCP_mcmc_period, "alpha[2]")
plot(sampCP_mcmc_period, "beta[1]")
plot(sampCP_mcmc_period, "beta[2]")
```

```{r}
sampCP_mcmc_period$DIC
loglik <- sampCP_mcmc_period$sims.list$LogLik # model with change point

waic <- waic(loglik)

waic
```

```{r}
modelAR <-"model {
  ## parameters: alpha,tau,m0
  #likeliohood 
  mu[1]<-Y[1]
  Yp[1]=mu[1]
  for (i in 2:N) {
    Y[i] ~ dnorm(mu[i],tau)
    mu[i]<-m0+alpha*Y[i-1]
    Yp[i] ~ dnorm(mu[i],tau) # prediction in sample
  }
  # prediction out of sample 
  ypOut[1] ~dnorm(m0+alpha*Y[N],tau) 
  for(k in 2:Npred){
  ypOut[k] ~dnorm(m0+alpha*ypOut[k-1],tau) 
  }
  sigma2<-1/tau
  #prior 
  alpha ~ dunif(-1.5,1.5)
  tau ~ dgamma(0.1, 10)
  m0 ~dnorm(0.0, 1.0E-10)
}"
```

```{r}
myData <- list(Y = JFK$series__raw, 
               N = length(JFK$series__raw),
               Npred = 12)  # quanti passi vuoi prevedere nel futuro

sampAR_mcmc <- jags(
  model.file = textConnection(modelAR), 
  data = myData,
  parameters.to.save = c("mu", "Yp", "ypOut", "alpha", "m0", "tau", "sigma2"),
  n.adapt = 1000, 
  n.iter = 5000, 
  n.chains = 2, 
  n.burnin = 2000
)

ssAR <- summary(sampAR_mcmc)
```

```{r}
#q1 <- ssAR["m0", c("2.5%")]
#q2 <- ssAR["m0", c("97.5%")]
#mu_mean <- ssAR["", muc("mean")]

plot(JFK$series__raw, col = "orange", type = "p", lwd = 2.0, xlim = c(200, 250), ylim = c(min(JFK$series__raw) - 0.5, max(JFK$series__raw) + 0.5), xlab = "time", ylab = "Y")
lines(c(1:468), ssAR[1:468, "2.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssAR[1:468, "97.5%"], type = "l", lty = 2, col = "blue")
lines(c(1:468), ssAR[469:936, "2.5%"], type = "l", lty = 2, col = "green")
lines(c(1:468), ssAR[469:936, "97.5%"], type = "l", lty = 2, col = "green")
points(ssAR[1:468, "mean"], col = "magenta", lwd = 1.0)
#abline(v = tmean, col = "red", lwd = 1.5)
#abline(v = q1, col = "red", lwd = 1.5, lty = 3)
#abline(v = q2, col = "red", lwd = 1.5, lty = 3)
labels <- c(
  paste0("Data"),
  paste0("Model"),
  paste0("Credible interval"),
  paste0("Predictive Interval"),
  paste0("Changepoint")
)
legend("topleft", legend = labels, col = c("orange", "magenta", "blue", "green", "red"), lty = 1, cex = 0.75)

# plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
# points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

# abline(v=tmean,col="red",lwd=1.5)
# abline(v=q1,col="red",lwd=1.5,lty = 3)
# abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampAR_mcmc, "m0")
plot(sampAR_mcmc, "alpha")
#plot(sampCP_mcmc_period, "mu[293]")
#plot(sampCP_mcmc_period, "xpred[1]")
#plot(sampCP_mcmc_period, "alpha[1]")
#plot(sampCP_mcmc_period, "alpha[2]")
#plot(sampCP_mcmc_period, "beta[1]")
#plot(sampCP_mcmc_period, "beta[2]")
```

```{}
```

```         
```
