---
title: "Final Project"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
editor_options: 
  markdown: 
    wrap: 72
---

# Final Project: JFK Passengers

**Source.**
<https://github.com/alan-turing-institute/TCPD/tree/master/datasets/jfk_passengers>

**Data.** The Port Authority collects monthly data on domestic and
international passengers, cargo, flights, and aircraft equipment types
from each carrier operating at PANYNJ airports.

**Task.** This dataset can be analyzed using a change-point model with
different means before and after the change point, possibly including a
linear drift after the change point. A basic model is
$$ y_t = \mu_t + \epsilon_t, $$ where
$$ \epsilon_t \sim {N}(0,\sigma_t^2), $$ and
$$ \mu_t = \mu_1,\quad \sigma_t = \sigma_1 \qquad \text{if } t \leq \tau, $$
while
$$ \mu_t = \mu_2,\quad \sigma_t = \sigma_2 \qquad \text{if } t > \tau. $$
Here $\tau$ is an unknown change-point parameter.

More complex change-point models may also be considered and discussed.
Time series models are encouraged; for example, one may include
different trends and different variances before and after the change
point.

```{r}
rm(list=ls())
JFK <- read.csv("JFK.csv") 
head(JFK)
tail(JFK)
dim(JFK)  
JFK$time__raw=as.Date(paste(JFK$time__raw,"-01",sep="")) 
plot(JFK$time__raw,JFK$series__raw,xlab="time",ylab="pass",main="JFK") 
plot(JFK$time__raw[1:13],JFK$series__raw[1:13],xlab="time",ylab="pass",main="JFK")
plot(JFK$time__raw[1:36],JFK$series__raw[1:36],xlab="time",ylab="pass",main="JFK")
plot(JFK$time__raw[400:468],JFK$series__raw[400:468],xlab="time",ylab="pass",main="JFK")
plot(JFK$time__raw[300:468],JFK$series__raw[300:468],xlab="time",ylab="pass",main="JFK")
plot(JFK$time__raw[275:400],JFK$series__raw[275:400],xlab="time",ylab="pass",main="JFK")

```

## Simple Changepoint Model

```{r}
mean(JFK$series__raw[1:350])
mean(JFK$series__raw[350:468])
var(JFK$series__raw)
length(JFK)
```

```{r}
library(rjags)
library(bayesplot)
library(jagsUI)
library(loo)
```

```{r}
modelChangePoint <-
"model{
  #
  for(t in 1:n){
    d[t]<-step(Tau-t)
  }
  # Likelihood
  for(i in 1:n){
    x[i]   ~ dnorm(mu[i],sinv)
    mu[i] <- m[J[i]] 
    J[i]<- d[i]+1 # move to (0,1) in (1,2)
    LogLik[i] <- log(dnorm( x[i],mu[i],sinv))  # this is for computing the WAIC, see later
    #predictive
    xpred[i] ~ dnorm(mu[i], sinv)
  }
  
  
  # Prior for d[i]
  Tau ~  dunif(1,n+10)
  # Prior for beta
  m[1] ~ dnorm(2500000,0.0000000000001)
  m[2] ~ dnorm(4000000,0.0000000000001)
  sinv ~  dunif(0,10)
  
}"
```

```{r}
myData=list(x=JFK$series__raw,n=length(JFK$series__raw)) # 

sampCP_mcmc <- jags(model.file=textConnection(modelChangePoint),data=myData,
                     parameters.to.save=c("mu","xpred","Tau","J","LogLik", "sinv"), 
                     n.adapt=1000, n.iter=10000,n.chains = 2,n.burnin = 2000)




ssCP=summary(sampCP_mcmc)

```

```{r}
q1=ssCP["Tau",c("2.5%")]
q2=ssCP["Tau",c("97.5%")]
tmean=ssCP["Tau",c("mean")]
mtrueT=c(rep(2354015,330),rep(4070000,168))

plot(mtrueT[1:468],col="black",type="l",lwd=2.0,ylim=c(min(JFK$series__raw)-0.5,max(JFK$series__raw)+0.5),xlab="time",ylab="Y")
lines(c(1:468),JFK$series__raw,type="p",col="orange")
lines(c(1:468),JFK$series__raw,type="l",col="orange")
abline(h=ssCP[1,"2.5%"],lty=2,col="blue",lwd=1.0)
abline(h=ssCP[1,"97.5%"],lty=2,col="blue",lwd=1.0)
abline(h=ssCP[468,"2.5%"],lty=2,col="blue",lwd=1.0)
abline(h=ssCP[468,"97.5%"],lty=2,col="blue",lwd=1.0)
abline(h=ssCP[469,"2.5%"],lty=2,col="pink",lwd=1.0)
abline(h=ssCP[469,"97.5%"],lty=2,col="pink",lwd=1.0)
abline(h=ssCP[936,"2.5%"],lty=2,col="green",lwd=1.0)
abline(h=ssCP[936,"97.5%"],lty=2,col="green",lwd=1.0)
points(ssCP[1:468,"mean"],col="magenta",lwd=2.0) 
abline(v=tmean,col="red",lwd=1.5)
abline(v=q1,col="red",lwd=1.5,lty = 3)
abline(v=q2,col="red",lwd=1.5,lty = 3)
labels=c(paste0('Data'),
           paste0('Model'),
           paste0('Credible interval'),
           paste0('Predictive Interval 1'),
           paste0('Predictive Interval 2'),
           paste0('Changepoint'))
legend("topleft", legend=labels, col=c("orange", "magenta","blue", "pink", "green", "red"), lty = 1, cex=0.75)

#plot(Ttrue-1,col="black",type="l",lwd=2.0,xlab="time",ylab="probability")
#points(1-(ssCP2[(TT+2):(2*TT+2)]-1),col="magenta") #

#abline(v=tmean,col="red",lwd=1.5)
#abline(v=q1,col="red",lwd=1.5,lty = 3)
#abline(v=q2,col="red",lwd=1.5,lty = 3)
plot(sampCP_mcmc,"Tau")
plot(sampCP_mcmc,'mu[1]')
plot(sampCP_mcmc,'mu[468]')
plot(sampCP_mcmc,'xpred[1]')
```

```         
```
